A thermodynamic view of dusty protoplanetary disks I: axisymmetric stability

### 2017-01-05 linear.tex
edits on the plane
new section 7 on secular GI. rest of paper might need adjusting. mostly doing non-SG, but one sec and one appendix has SG

\section{Linearized equations}\label{linear_problem}
To compute explicit solutions to the linear problem, 
we consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta f (r, z) = \delta f_1(r,z)\exp{(\ii k_x r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k_x$ is a real wavenumber such that $|k_xr|\gg 1$, and the 
amplitude $\dd f_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k_x$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k_x>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

@@

\begin{align}
  \ii\sigma W &= \ii k_x \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k_x Q,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right]  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k_x \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy} 
\end{align}  
where $^\prime \equiv \p_z$ and recall $\bm{F} \equiv -\nabla P/\rho$. 
The linearized dust-duffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
Note that we have assumed a temperature profile that only depends on $r$.

@@

Eq. \ref{lin_mass}---\ref{lin_energy} is a set of ordinary
differntial equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained. We now discuss solutions to these equations in an unstratified disk 
in \S\ref{si} and a stratified disk in \S\ref{results}.

@@

\section{Streaming instability}\label{si}
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain an algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

@@

>>>>

\section{Linearized equations}\label{linear_problem}
To compute explicit solutions to the linear problem, 
we consider radially-localized axisymmetric disturbances of the form  
\begin{align}
  \delta X (r, z) = \delta X_1(r,z)\exp{(\ii k_x r)},
\end{align} 
%and similarly for $\dd P$ and $\dd\bm{v}$. 
where $k_x$ is a real wavenumber such that $|k_xr|\gg 1$, and the 
amplitude $\dd X_1(r,z)$ is 
a slowly-varying function of $r$. Then 
$\p_r\to i k_x$ when acting on the above primitive perturbations, and we may
neglect curvature terms. We take  
$k_x>0$ without loss of generality. Hereafter, we drop the subscript 1
on the amplitudes. 
%The frequency $\sigma = \omega +
%\ii s$ is generally complex, with $\omega$ being the real frequency
%and $s$ is the real growth rate.

@@

\begin{align}
  \ii\sigma W &= \ii k_x \dd v_r + \dd v_z^\prime +
  \dd v_r \p_r\ln{\rho} + \dd v_z\p_z\ln{\rho},\label{lin_mass}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi 
% +  \delta\bm{F}\cdot\hat{\bm{r}}
- W F_r - \ii k_x Q - \ii k_x\dd\psi,\label{lin_xmom}\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z, \label{lin_ymom}\\
  -\ii\sigma\dd v_z &= - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right] - \dd\psi^\prime  %\delta\bm{F}\cdot\hat{\bm{z}} 
,\label{lin_zmom}\\
  \ii\sigma Q &= \frac{P}{\rho}\left(\ii k_x \dd v_r + \dd
               v_z^\prime\right) + \frac{1}{\rho}\left(\dd v_r\p_rP + \dd v_z \p_zP\right)\notag\\
                &\phantom{=}-\frac{P}{\rho} \dd v_r\p_r
               \ln{c_s^2} %, \label{lin_energy} 
%+ \dd v_z \p_z\ln{c_s^2}\right),\label{lin_energy} 
               - \frac{\dd\mathcal{C}}{\rho},\label{lin_energy}\\
\dd\psi^{\prime\prime}  &= 4\pi G \rho W + k_x^2\dd\psi, \label{lin_sg}
\end{align}  
where $^\prime \equiv \p_z$ and recall $\bm{F} \equiv -\nabla P/\rho$. 
The linearized dust-duffusion function 
$\dd\mathcal{C}$ is given in  Appendix \ref{lin_dust}. 
% and $\dd\bm{F}$ is the linearized pressure
%force, given in Appendix \ref{lin_press}. 
Note that we have assumed a temperature profile that only depends on $r$.

@@

Eq. \ref{lin_mass}---\ref{lin_sg} is a set of ordinary
differntial equations in $z$. All coefficients and amplitudes are
evaluated at a fiducial radius $r=r_0$, but their full $z$-dependence
is retained. We now discuss solutions to these equations for a
razor-thin, self-gravitating disk in \S\ref{sgi}; and for 3D
non-self-gravitating, unstratified and stratified disks 
in \S\ref{si} and \S\ref{results}, respectively.

@@

\section{Secular gravitational instability}\label{sgi}
We first consider a razor-thin disk so that $\rho =
\Sigma\delta(z)$. Here $\delta$ is the delta function and $\Sigma$ is
the total surface density. Note that $\epsilon$ and 
$\tepsilon$ now refers to the global dust-to-gas ratio and dust 
fraction, respectively. We neglect the vertical dimension   
in Eq. \ref{lin_mass}---\ref{lin_energy}. The background disk is
uniform. The linearized dust-gas drag term is then
$-\delta\mathcal{C}/\rho = \tstop\tepsilon c_s^2k_x^2 Q$. The
thin-disk solution to Eq. \ref{lin_sg} is $\dd\psi = -2\pi G
\Sigma W/\left|k_x\right|$

These simplifications yield the dispersion relation
\begin{align*}
  \left(\ii\sigma - \tstop\tepsilon c_s^2k_x^2\right)\left( 2 \pi G
    \Sigma \left|k_x\right|  - \kappa^2 + \sigma^2 \right) = \ii
  \sigma c_s^2 k_x^2\left(1 - \tepsilon\right). 
\end{align*}
Searching for slowly and purely growing modes, $\sigma = \ii s$ with 
$|s|\ll \kappa$, we find 
\begin{align}  
s = \frac{\tstop\tepsilon c_s^2 k_x^2 \left( 2 \pi \Sigma G
    \left|k_x\right| - \kappa^2\right)}{\kappa^2 -2 \pi \Sigma G
    \left|k_x\right| + c_s^2k_x^2\left(1-\tepsilon\right) }. \label{sgi_disp}
\end{align}
This is just the secular gravitational instability in the absence of
turbulent dust diffusion but strong drag
\citep[][ their Eq. 13 becomes Eq. \ref{sgi_disp} in this limit with
a change of variables]{takahashi14}. This is gravitational instability
mediated by dust-gas drag. A similar effect occurs in viscous
self-gravitating gas disks \citep{gammie96,lin16}. In fact, if we
identify $\nu \equiv \tstop c_s^2$ as a kinematic viscosity, then
Eq. \ref{sgi_disp} is effectively 
\citeauthor{gammie96}'s Eq. 18.

This exercise shows that the one-fluid framework, futher simplified by
the terminal velocity approximation, is sufficient to capture the SGI
in the strong drag limit.

\section{Streaming instability}\label{si}
We now consider 3D but non-self-gravitating disks.  
In the special case of an unstratified disk (neglecting the vertical
component of the stellar gravity), we may also Fourier analyze in $z$
to obtain an algebraic dispersion relation of the form  
$\sum_{j=0}^{5}c_j(k_x,k_z)\sigma^j = 0$, where $k_z$ is a real
vertical wavenumber. The coefficients $c_j$ can be read 
off Eq. \ref{streaming_dispersion} in  Appendix \ref{compressible_streaming}. 
There we also show that this dispersion relation reduces to that for
the streaming instability (SI) in the limit of incompressible gas and small
$\tstop$ \citep{youdin05a,jacquet11}.

@@

% \subsection{Eulerian interpretation}

% We can also consider SI in the Eulerian sense \citep[as
% did][]{jacquet11}. Eq. \ref{thermal_instability} imply SI requires
% \begin{align*}
%   \frac{\imag\left[\delta\mathcal{C}\nabla\cdot\dd\bm{v}^*\right]}{\real(\sigma)}<0. 
% \end{align*}
% The linearized dust diffusion or `cooling' term $\delta\mathcal{C}$
% for this problem is given by Eq. \ref{lin_drag_si}. Thus the
% requirement is 
% \begin{align*}
% \tepsilon |\bm{k}|^2 
%   \frac{\imag\left(\ii \sigma^* QW^*\right)}{\real(\sigma)}
% -k_x F_r\left( 1-2\tepsilon\right) 
%  | W |^2\frac{\imag\left(\sigma^*\right)}{\real(\sigma)} < 0, 
% \end{align*}
% where we have used Eq. \ref{streaming_mass}. 
% For slow growth and/or large $\left|\bm{k}\right|^2$ the second term 
% is small compared to the first. In this case instablity is due to

### 2017-01-05 model.tex
new section 7 on secular GI. rest of paper might need adjusting. mostly doing non-SG, but one sec and one appendix has SG

In the limit of strong dust-gas drag, $\tstop\to0$, the mixture obeys 
the first-order (in $\tstop$) single-fluid equations: 
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H}_\mathrm{drag} + \mathcal{H}_\mathrm{eff}  - \Lambda  \label{tempeq} 
\end{align}
\citep{laibe14}, where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}. 
\end{align}
For an ideal gas the pressure is given
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

@@

For the disk problem the external potential $\Phi$ is  
that for a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
where $G$ is the gravitational constant, and $(r,\phi, z)$ are
cylindrical co-ordinates centered on the star. 
The second equality is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria.

>>>>

In the limit of strong dust-gas drag, $\tstop\to0$, the mixture obeys 
the first-order (in $\tstop$) single-fluid equations: 
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi_\mathrm{tot} - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H}_\mathrm{drag} + \mathcal{H}_\mathrm{eff}  - \Lambda  \label{tempeq} 
\end{align}
\citep{laibe14}, where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}. 
\end{align}
For an ideal gas the pressure is given
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

@@

The total gravitational potential $\Phi_\mathrm{tot}=\Phi + \psi$,
where $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
and $\psi$ is the disk potential satisfying
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.
\end{align}
Here, $G$ is the gravitational constant, and $(r,\phi, z)$ are 
cylindrical co-ordinates centered on the star.  
The second equality in Eq. \ref{thin_disk_potential} is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{eqm}).

We include the gravitational potential of the mixture 
$\psi$ for completeness but will neglect it in most of this
work. Henceforth we consider non-self-gravitating disks ($\psi \equiv
0$) unless explicitly stated otherwise.

### 2017-01-05 paper.tex
new section 7 on secular GI. rest of paper might need adjusting. mostly doing non-SG, but one sec and one appendix has SG

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{chiang08,barranco09,lee10}. These lead to structure and
planetesimal formation, as well as turbulence. 
More recently, new 
dusty instabilities have also been uncovered through numerical
simulations \citep{loren15,loren16,lamb16}, but these are not yet well
understood.

@@

We begin our formal discussion in \S\ref{setup} by listing the 
basic equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} that are used in
explicit calculations. In \S\ref{limits} we discuss 
general stability properties of dusty disks. We then move onto
numerical stability analyses. In  
\S\ref{linear_problem} we list the linearized equations, and solve 
them for the SI in \S\ref{si}, and in 
 \S\ref{results} to study the dusty VSI. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

>>>>

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep[SGI,][]{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{chiang08,barranco09,lee10}. These lead to structure and
planetesimal formation, as well as turbulence. 
More recently, new 
dusty instabilities have also been uncovered through numerical
simulations \citep{loren15,loren16,lamb16}, but these are not yet well
understood.

@@

We begin our formal discussion in \S\ref{setup} by listing the 
basic equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} that are used in
explicit calculations. In \S\ref{limits} we discuss 
general stability properties of dusty disks. We then move onto
explicit stability analyses. In  
\S\ref{linear_problem} we list the linearized equations. We solve them 
to recover SGI in \S\ref{sgi}, to interpret the SI in \S\ref{si}, and
in  \S\ref{results} to extend the VSI to dusty disks. We  
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.
